A comparison on R tools for nonlinear least squares modeling

### 2023-05-02 RefactoringNLS/ComparingNLStools.Rmd

In a Google Summer of Code project "Improvements to `nls()`",
we considered the features and limitations of the R function `nls()`
in the context of improving and rationalizing R tools for nonlinear regression.
The rich features of this function are weakened by a number of deficiencies and
inconsistencies, in particular a lack of stabilization of the Gauss-Newton solver.
Further considerations are the usability and  maintainability of
the code base that provides the functionality `nls()` claims to offer.
Various packages, including our `nlsr`, provide alternative capabilities.
This article considers the differences in goals, approach and features of
different tools for nonlinear least squares modeling in R. Discussion of 
these matters is relevant to improving R generally as well as its 
nonlinear estimation tools.

@@

`nls()` is the tool in base R, the primary software software distribution
from the Comprehensive R Archive Network (*https://cran.r-project.org*),
for estimating nonlinear statistical models. The function dates to the 1980s and the 
work related to @bateswatts in S (see 
*https://en.wikipedia.org/wiki/S_%28programming_language%29*).

@@

- handles formulas that include R functions, even ones which call calculations in
  other programming languages
- allows data to be weighted or subset
- can estimate bound constrained parameters
- provides a mechanism for handling partially linear models
- permits parameters to be indexed over a set of related data
- produces measures of variability (i.e., standard error estimates) for the 
estimated parameters
- has related profiling capabilities for exploring the likelihood surface as 
parameters are changed
- links to a number of pre-coded (`selfStart`) models that do not require
  initial parameter values.

@@

The tools we will consider are the base-R `nls()` function 
and a number of packages that are available on the CRAN repository.
Of these, we will pay particular attention to **nlsr** (@nlsr2023manual),
**minpack.lm** (@minpacklm12), **gslnls** (@gslnls23), and **nls2** (@nls2-22).

There are a number of other CRAN packages worth commentary, and we will give
some capsule comments later. It is likely that there are a number of other 
tools that could be relevant to
this comparison in the Bioconductor (@Bioconductor) collection 
and on repositories like
Github (https://github.com) and Gitlab (https://about.gitlab.com), but we will 
not pursue those possibilities here.

@@

R tools are our interest, but they often call programs in other programming
languages. There are advantages in using a well-established program that has
been much used and tested. Early versions of R were relatively sluggish in 
computational speed. The downside is that we must interface between two or more
programming languages, adding to the risk of errors. Changes to one or the 
other language or its library support may
force maintenance issues. Moreover, the pool of expertise with languages like C
and Fortran is diminishing.

@@

Most researchers are challenged to program the differential calculus 
required to specify the Jacobian to a nonlinear estimation tool. 
The work-around is generally to use a numerical approximation,
but there is the possibility of using symbolic or automatic differentiation.
There are also particular models where workers have provided code for the
Jacobian.

@@

Reality imposes conditions on parameters. While general constraints can
be difficult to include, bounds on parameters can be accommodated by many 
solvers. For some reason, fixed parameters (masks) are not often considered,
but are convenient when a parameter has a commonly accepted value that we
later wish to allow to be estimated from the data.

@@

While the aim of nonlinear estimation is to return a set of parameters for a
particular model, there is a variety of additional information which may be
provided. From the perspective of the nonlinear least squares problem, there
are measures of work and information on the quality of the solution, while
for statistical purposes measures of uncertainty in the parameters are helpful.

Some tools allow constraints to be imposed on the solution, and it is helpful
to know if these constraints are active at the proposed solution.

@@

We believe that the availability of good documentation with examples that users
can modify to deal with their own problems is essential. The presence of vignettes, 
demos, and tests indicates a package is well-supported.

@@

`onls` (@R-onls-manual) -- The usual optimization in estimating nonlinear models
is the vertical difference between the dependent variable and the functional
model. `onls` minimizes the sum of **orthogonal** residuals. The objective is
therefore different, and involves nontrivial extra calculation. A vignette with
the package and the blog article 
https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/ 
give some description with illustrative graphs. `onls` appears to be limited to 
problems with one independent and one dependent variable. The Wikipedia article
https://en.wikipedia.org/wiki/Total_least_squares presents an overview of some
ideas, with references to the literature. Our view is that the approach needs wider 
discussion and tutorial examples to allow its merits to be judged.

@@

`nlsic` (@nlsic-manual) -- This solves non linear least squares problems with optional
equality and/or inequality constraints. It is clearly **not** about modeling, and
the input and output are quite different from class `nls` methods.

@@

`nlsmsn` (@m-nlsmsn) -- Fit univariate non-linear scale mixture of skew-normal(NL-SMSN) 
regression, with details in Garay, Lachos and Abanto-Valle (2011). The problem 
here is to minimize an objective which is modified from the traditional sum of squared
residuals.

@@

`nls2` (@nls2-22) -- Nonlinear least squares by brute force has some similar
motivations, but uses `nls()` within multiple trials. The package and its
author have a long history in the R community.

@@

The Hobbs weed infestation problem (@cnm79 [page 120]) is a growth curve modeling 
task which is seemingly straightforward but turns out to be quite nasty. Because
this problem has a very succinct statement, it often provides the "short 
reproducible example" much beseeched of contributors to R mailing lists. 
Nevertheless, this is not an artificially created test problem, but one that
came from a field researcher.

The data and a graph of it is given below.

@@

$$  H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}$$
Let us use $D_{i,j}$ for the elements of the second term of this expression.
What is generally called **Newton's method** for function minimization tries
to set the gradient to zero (to find a stationary point of the function $S(p)$).
This leads to the **Newton's equation**

@@

In reality, there are many problems were $D$ should not be ignored, but the work
to compute it precisely is considerable. A number of work-arounds have been 
proposed, of which the Levenberg-Marquardt (@Levenberg1944, @Marquardt1963) is
the most commonly used. Henceforth we will simply say "Marquardt" as we believe
he was the first to actually incorporate the ideas into a practical computer
program.

@@

Let us specify in R the three model formulae and set some starting
values for parameters. These starting points are NOT equivalent, and
are deliberately crude choices. Workers performing many calculations 
of a similar nature should try to provide good starting points to reduce
computation time and the chance a false solution is found.

@@

One of the useful features of `nls()` is the possibility of a `selfStart` model,
where starting parameter values are not required. We discuss aspects of this
feature at several points below. However, in the event that a `selfStart` model 
is not available, `nls()` sets all the starting parameters 
to 1. This is, in our view, tolerable, but could be improved by using a set of values
that are all slightly different, which, in the case of the example 
model $y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)$
would avoid a singular Jacobian because $b$ and $d$ were equal in value. 
Program modifications to give a sequence like  1.0, 1.1, 1.2, 1.3 for the four 
parameters are fairly obvious.

@@

```{r ex02minpack}
library(minpack.lm)
unlm1<-try(nlsLM(formula=frmu, start=stu1, data=weeddf))
summary(unlm1) # Use summary() to get display
# unlm1 # Note the difference. Use this form to get sum of squares
# pnlm0(unlm1)  # Short form of output
snlm1<-try(nlsLM(formula=frms, start=sts1, data=weeddf))
pnlm0(snlm1)
# summary(snlm1)
## Following structure needed to deal with M1mac issue
tnlm1<-try(nlsLM(formula=frmt, start=stt1, data=weeddf))
if (inherits(tnlm1, "try-error")) {
   cat("Failure to compute solution -- likely singular Jacobian\n")
} else {  
   pnlm0(tnlm1) # short form to give sum of squares, else use summary(tnlm1)
}   
```

@@

- `nlsLM()` includes bounds in the standard call, but we have observed cases
   where it fails to get the correct answer. From examination of the code,
   we believe the authors have not taken into account all possibilities, though
   it should be noted that all programs have some weakness in
   regard to constrained optimization. Software creators have to work with
   assumptions on the extremity of scale that they are willing to countenance,
   and sometimes problems will be outside the scope envisaged.

@@

While fixed weights may be useful, there are a number of problems for which
we want weights that are determined at least partially from the model 
parameters, for
example, a measure of the standard deviation of observations.
Such dynamic weighting situations are discussed in the vignette 
"Introduction to nlsr" of package `nlsr` in section 
*Weights that are functions of the model parameters* and 
`minpack.lm` offers a function `wfct()` to 
facilitate such weighting. Care is advised
in applying such ideas.

@@

As the `nls()` **man** page states, when the "port" algorithm is used with the
`trace` argument TRUE, the iterations display the objective function value which 
is 1/2 the sum of squares (or deviance). It is likely that the trace display is
embedded in the Fortran of the `nlminb` routine that is called to execute 
the "port" algorithm, but the factor of 2 discrepancy is nonetheless
unfortunate for users.

### Failure to return best result achieved

>>>>

In a Google Summer of Code project "Improvements to `nls()`",
we considered the features and limitations of the R function `nls()`
in the context of improving and rationalizing R tools for nonlinear regression.
The rich features of this function are weakened by several deficiencies and
inconsistencies, in particular a lack of stabilization of the Gauss-Newton solver.
Further considerations are the usability and  maintainability of
the code base that provides the functionality `nls()` claims to offer.
Various packages, including our `nlsr`, provide alternative capabilities.
This article considers the differences in goals, approaches, and features of
different tools for nonlinear least squares modeling in R. Discussion of 
these matters is relevant to improving R generally as well as its 
nonlinear estimation tools.

@@

`nls()` is the tool in base R, the primary software distribution
from the Comprehensive R Archive Network (*https://cran.r-project.org*),
for estimating nonlinear statistical models. The function dates to the 1980s and the 
work related to @bateswatts in S (see 
*https://en.wikipedia.org/wiki/S_%28programming_language%29*).

@@

- handles formulas that include R functions, even ones that call calculations in
  other programming languages
- allows data to be weighted or subset
- can estimate bound-constrained parameters
- provides a mechanism for handling partially linear models
- permits parameters to be indexed over a set of related data
- produces measures of variability (i.e., standard error estimates) for the 
estimated parameters
- has related profiling capabilities for exploring the likelihood surface as 
parameters are changed
- links to many pre-coded (`selfStart`) models that do not require
  initial parameter values.

@@

The tools we will consider are the base-R `nls()` function 
and some packages that are available on the CRAN repository.
Of these, we will pay particular attention to **nlsr** (@nlsr2023manual),
**minpack.lm** (@minpacklm12), **gslnls** (@gslnls23), and **nls2** (@nls2-22).

There are some other CRAN packages worth commentary, and we will give
some capsule comments later. There are likely several other 
tools that could be relevant to
this comparison in the Bioconductor (@Bioconductor) collection 
and on repositories like
GitHub (https://github.com) and Gitlab (https://about.gitlab.com), but we will 
not pursue those possibilities here.

@@

R tools are our interest, but they often call programs in other programming
languages. There are advantages in using a well-established program that has
been used and tested. Early versions of R were relatively sluggish in 
computational speed. The downside is that we must interface between two or more
programming languages, adding to the risk of errors. Changes to one or the 
other language or its library support may
force maintenance issues. Moreover, the pool of expertise with languages like C
and Fortran is diminishing.

@@

Most researchers are challenged to program the differential calculus 
required to specify the Jacobian to a nonlinear estimation tool. 
The workaround is generally to use a numerical approximation,
but there is the possibility of using symbolic or automatic differentiation.
There are also particular models where workers have provided code for the
Jacobian.

@@

Reality imposes conditions on parameters. While general constraints can
be difficult to include, bounds on parameters can be accommodated by many 
solvers. For some reason, fixed parameters (masks) are not often considered
but are convenient when a parameter has a commonly accepted value that we
later wish to allow to be estimated from the data.

@@

While nonlinear estimation aims to return a set of parameters for a
particular model, there is a variety of additional information which may be
provided. From the perspective of the nonlinear least squares problem, there
are measures of work and information on the quality of the solution, while
for statistical purposes measures of uncertainty in the parameters are helpful.

Some tools allow constraints to be imposed on the solution, and it is helpful
to know if these constraints are active in the proposed solution.

@@

We believe that the availability of good documentation with examples that users
can modify to deal with their problems is essential. The presence of vignettes, 
demos, and tests indicates a package is well-supported.

@@

`onls` (@R-onls-manual) -- The usual optimization in estimating nonlinear models
is the vertical difference between the dependent variable and the functional
model. `onls` minimizes the sum of **orthogonal** residuals. The objective is
therefore different and involves nontrivial extra calculation. A vignette with
the package and the blog article 
https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/ 
give some description with illustrative graphs. `onls` appears to be limited to 
problems with one independent and one dependent variable. The Wikipedia article
https://en.wikipedia.org/wiki/Total_least_squares presents an overview of some
ideas, with references to the literature. Our view is that the approach needs wider 
discussion and tutorial examples to allow its merits to be judged.

@@

`nlsic` (@nlsic-manual) -- This solves nonlinear least squares problems with optional
equality and/or inequality constraints. It is clearly **not** about modeling, and
the input and output are quite different from class `nls` methods.

@@

`nlsmsn` (@m-nlsmsn) -- Fit univariate non-linear scale mixture of skew-normal(NL-SMSN) 
regression, with details in Garay, Lachos, and Abanto-Valle (2011). The problem 
here is to minimize an objective that is modified from the traditional sum of squared
residuals.

@@

`nls2` (@nls2-22) -- Nonlinear least squares by brute force has some similar
motivation, but uses `nls()` within multiple trials. The package and its
author have a long history in the R community.

@@

The Hobbs weed infestation problem (@cnm79 [page 120]) is a growth curve modeling 
task that is seemingly straightforward but turns out to be quite nasty. Because
this problem has a very succinct statement, it often provides the "short 
reproducible example" much beseeched of contributors to R mailing lists. 
Nevertheless, this is not an artificially created test problem, but one that
came from a field researcher.

The data and a graph of it are given below.

@@

$$  H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}$$
Let us use $D_{i,j}$ for the elements of the second term of this expression.
What is generally called **Newton's method** for function minimization tries
to set the gradient to zero (to find a stationary point of the function $S(p)$).
This leads to **Newton's equation**

@@

In reality, there are many problems were $D$ should not be ignored, but the work
to compute it precisely is considerable. Many workarounds have been 
proposed, of which the Levenberg-Marquardt (@Levenberg1944, @Marquardt1963) is
the most commonly used. Henceforth, we will simply say "Marquardt", as we believe,
was the first to actually incorporate the ideas into a practical computer
program.

@@

Let us specify in R the three model formulae and set some starting
values for parameters. These starting points are NOT equivalent and
are deliberately crude choices. Workers performing many calculations 
of a similar nature should try to provide good starting points to reduce
computation time and the chance a false solution is found.

@@

One of the useful features of `nls()` is the possibility of a `selfStart` model,
where starting parameter values are not required. We discuss aspects of this
feature at several points below. However, if a `selfStart` model 
is not available, `nls()` sets all the starting parameters 
to 1. This is, in our view, tolerable, but could be improved by using a set of values
that are all slightly different, which, in the case of the example 
model $y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)$
would avoid a singular Jacobian because $b$ and $d$ were equal in value. 
Program modifications to give a sequence like  1.0, 1.1, 1.2, 1.3 for the four 
parameters are fairly obvious.

@@

```{r ex02minpack}
library(minpack.lm)
unlm1<-try(nlsLM(formula=frmu, start=stu1, data=weeddf))
faraway::sumary(unlm1) # Use summary() to get display
# unlm1 # Note the difference. Use this form to get sum of squares
# pnlm0(unlm1)  # Short form of output
snlm1<-try(nlsLM(formula=frms, start=sts1, data=weeddf))
pnlm0(snlm1)
# summary(snlm1)
## Following structure needed to deal with M1mac issue
tnlm1<-try(nlsLM(formula=frmt, start=stt1, data=weeddf))
if (inherits(tnlm1, "try-error")) {
   cat("Failure to compute solution -- likely singular Jacobian\n")
} else {  
   pnlm0(tnlm1) # short form to give sum of squares, else use summary(tnlm1)
}   
```

@@

- `nlsLM()` includes bounds in the standard call, but we have observed cases
   where it fails to get the correct answer. From examination of the code,
   we believe the authors have not taken into account all possibilities, though
   it should be noted that all programs have some weakness regarding constrained
   optimization. Software creators have to work with
   assumptions on the extremity of scale that they are willing to countenance,
   and sometimes problems will be outside the scope envisaged.

@@

While fixed weights may be useful, there are many problems for which
we want weights that are determined at least partially from the model 
parameters, for
example, a measure of the standard deviation of observations.
Such dynamic weighting situations are discussed in the vignette 
"Introduction to nlsr" of package `nlsr` in section 
*Weights that are functions of the model parameters* and 
`minpack.lm` offers a function `wfct()` to 
facilitate such weighting. Care is advised
in applying such ideas.

@@

As the `nls()` **man** page states, when the "port" algorithm is used with the
`trace` argument TRUE, the iterations display the objective function value which 
is 1/2 the sum of squares (or deviance). The trace display is likely
embedded in the Fortran of the `nlminb` routine that is called to execute 
the "port" algorithm, but the factor of 2 discrepancy is nonetheless
unfortunate for users.

### Failure to return the best result achieved
